---
title: "Stat184Final"
author: "Patrick Erickson, Aiden Shindel, Jordan Brophy"
format: pdf
editor: visual
csl: mla.csl
bibliography: references.bib
---

# Logistic Regression Analysis for Basketball Teams: Can we predict whether one team will win over another?

## Introduction

In basketball, a number of statistics can be used in order to predict the winner of a game. For instance a teams field goal percentage often directly coorelates to how many points they score in a game. For our project we decided to create a model that creates a prediction on which team will win in a head to head matchup based on variables such as date of the game, field goal percentage, points, opponent points, location, and who they played. The date repesnets when the game was played. A win or a loss prediction can fluxuate based on when the game was played due to teams suffering from injuries and other possible confounding variables. Field goal percentage is calculated by the the number of field goals made divided by the number of field goals attempted. A field goal is any shot that is taken by a team during the duration of the game not including free throws. Often times teams will shoot and make anywhere between 40% and 60% of there shots throughout a game. The points variable represents the number of points a team scores in a game. The winner of the game is the team that scores the most points. Opponent points is simalar to points but rather the amount of points your team allows the opponent team to score. Location is where the game is played and reprented by either a home game or an away game. Often times a home game is better for a team because of the fan engagment which can help a teams morale and encourage better play. Who the team plays against represents the opponent for the match. Different opponents will be of different skill levels a better team may be more difficult or easier to beat depending on who it is.

This data is findable. It was located on Kaggle which is a hub for data science and machine learning that provides data sets that are easily

```{r}
library(tidyr)
library(ggplot2)
library(dplyr)
```

You can add options to executable code like this

```{r}

```

## How can we predict whether one team will win against another?
