---
title: "Stat184Final"
author: "Patrick Erickson, Aiden Shindel, Jordan Brophy"
format: pdf
editor: visual
csl: mla.csl
bibliography: references.bib
---

# Logistic Regression Analysis for Basketball Teams: Can we predict whether one team will win over another?

## Introduction

  In basketball, a number of statistics can be used in order to predict the winner of a game. For instance a teams field goal percentage often directly coorelates to how many points they score in a game. For our project we decided to create a model that creates a prediction on which team will win in a head to head match up based on variables such as date of the game, field goal percentage, points, opponent points, location, and who they played. The date represents when the game was played. A win or a loss prediction can fluctuate based on when the game was played due to teams suffering from injuries and other possible confounding variables. Field goal percentage is calculated by the the number of field goals made divided by the number of field goals attempted. A field goal is any shot that is taken by a team during the duration of the game not including free throws. Often times teams will shoot and make anywhere between 40% and 60% of there shots throughout a game. The points variable represents the number of points a team scores in a game. The winner of the game is the team that scores the most points.  Opponent points is similar to points but rather the amount of points your team allows the opponent team to score. Location is where the game is played and represented by either a home game or an away game. Often times a home game is better for a team because of the fan engagement which can help a teams morale and encourage better play. Who the team plays against represents the opponent for the match. Different opponents will be of different skill levels a better team may be more difficult or easier to beat depending on who it is. 

This data follows the Fair principle. This data is find able. It was located on Kaggle which is a hub for data science and machine learning that provides data sets that are easily. The data was accessible because Kaggle is a free to use site once you make an account. This data is also available on other sites but a Kaggle author just collected it and made it easily available. The data is interpenetrate because the data set has clearly defined variables and is not missing any data for the year in which the data was collected. The data is also reusable because it provides many interesting insights for the games played during the year in which it was collected. 
The data also follows the Care principles. It provides collective benefit because it is a Data set that easily and properly shows match up data that any team can use to scout and prepare to face another team. The teams that are involved in this data set also have the authority to control it because it is their teams that are being represented. Any team that is represented in this data set also  have the ability to use this data set for there own collective benefit. This data set is used for the purpose of benefiting the teams involved and getting a better understanding of each individual match up. 

```{r}
library(tidyr)
library(ggplot2)
library(dplyr)
```

You can add options to executable code like this

```{r}

```

The `echo: false` option disables the printing of code (only output is displayed).
